Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-09 paper2.qmd
section:appendix; process:write

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Causal analysis
  - Directed Acyclic Graphs
  - Bayesian statistical methods
  - Thurstonian model
  - Comparative judgement
  - Probability
  - Statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
---

@@

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed. The present study primarily focuses on two: the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The following sections begin with a brief overview of Thurstone's theory and a detailed examination of these issues. Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

This section examines the two critical issues in the CJ literature that serve as the primary focus of this study. The first is the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as the measurement model in CJ assessments. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to examine the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.

@@

# An updated theoretical and statistical model for CJ {#sec-theory}

This section presents a theoretical model for CJ that extends Thurstone's theory. It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory. The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.

@@

<!-- []{style="color:red;"} -->

@@

@@

@@

<!-- Notably, the theory assumes the judge compares each pair of stimuli a "sufficient number of times" to accurately calculate the relative separation probability [@Thurstone_1927b, pp. 267]. -->

<!-- It is assumed that the single observer compares each pair of stimuli sufficient number of times so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp. 267] -->

@@

@@

<!-- , as well as when eliminating data through ad hoc univariate procedures -->
<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->

@@

# Declarations {.appendix .unnumbered}

@@

# References {.unnumbered}

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
---

@@

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across different stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level. For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on the validity of CJ scores indicates their capacity to represent the traits under measurement accurately [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Moreover, research on CJ's practical applicability highlights its versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
Nevertheless, despite the increasing number of CJ studies, the prevalence of unsystematic and fragmented research approaches has left several critical issues unaddressed. The present study primarily focuses on two issues: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The following sections begin with a brief overview of Thurstone's theory followed by a detailed examination of these issues. Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, given that the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

This section examines the two critical issues in the CJ literature that serve as the primary focus of the present study. The first is related to the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second concern with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as a measurement model in CJ assessments. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating writing quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying level of writing quality (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.

@@

# Updating CJ's theoretical and statistical model {#sec-theory}

This section presents a theoretical model for CJ that extends Thurstone's theory. It uses causal inference and, in particular, the structural approach where structural causal models (SCMs) and directed acyclic graphs (DAGs) articulate the core theoretical principles of CJ theory. The model also incorporates several assessment design features that influence CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition, the study uses Bayesian statistical methods to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.

@@

### The population model {#sec-theory-theoretical_P}
<!-- []{style="color:red;"} -->

@@

structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]

causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]

<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->

<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests. Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp. 4]. -->

<!-- The main reason to focus on identification as a first step is that if a parameter cannot be identified using “infinite” and error free data, we certainly cannot learn anything about it using finite data.1 So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp. 5] -->

Assuming population data or more commonly known as census data, we ...

@@

### The assessment design features {#sec-theory-theoretical_AD}

@@

<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp. 267] -->

@@

<!-- Statements about the real world from observed data require assumptions. This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg. The validity of our statements therefore critically hinges on the validity of those assumptions. If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are [@Kohler_et_al_2019, pp. 150] -->

@@

@@

# Declarations {.unnumbered .appendix appendix-style=plain}

@@

# Appendix {#sec-appendix .appendix appendix-style=plain}

This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in @sec-theory. It does not, however, offer a comprehensive overview of statistical and causal inference methods. Readers seeking more in-depth understanding may wish to explore introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find it helpful to consult introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020. For more advanced study, readers may refer to seminal intermediate papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020.

## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}

<!-- 1. What does empirical research do first? -->
Empirical research uses evidence from observation and experimentation to address real-world challenges. In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010]. For instance, researchers might be interested in answering the following question: “To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$. Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability. In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value.

<!-- 2. What do they do next? -->
Researchers would then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010]. The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.

![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}

<!-- 3. But, what do we need to produce a good estimate? -->
While numerous methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006]. If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i}| T_{i}=1]$ and $E[Y_{i}| T_{i}=2]$, respectively. The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.

<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met. Usually these estimates are expressed in terms of the proportions of groups  -->

<!-- 4. But we want these estimands to be causal -->
However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply identify associations. In the example, researchers can interpret the associational estimate represented by the Z-statistic as causal. This interpretation relies on the data meeting the assumptions of the Z-test and the data being collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. This recognition stems from their ability to enable researchers to interpret associational estimates as causal. They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010]. Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from other variables $X$ and the outcome $Y$ [@Morgan_et_al_2014; @Neal_2020].

<!-- 5. Experiments are not available to everyone, then causal inference comes to the rescue -->
Nevertheless, researchers often face constraints that limit their ability to conduct randomized experiments. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these cases, causal inference offers a valuable alternative for generating causal estimates and understanding the mechanisms underlying specific data. In addition, the framework can provide significant theoretical insights that can enhance the design of experimental and observational studies [@McElreath_2020].

## Empirical research under causal inference {#sec-appendix-B .appendix appendix-style=plain}

<!-- 1. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, the *causal inference* framework is designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in understanding and defining causal effects.

<!-- 2. But how does it work? -->
*Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. Although a comprehensive discussion of causes and counterfactuals is beyond the scope of this document, it is possible to provide a brief overview of how the framework addresses the fundamental problem of causality. Using the example outlined in @sec-appendix-B, the framework begins by defining the *individual causal effect* (ICE) as the difference between the students' potential outcomes: $\tau_{i} = Y^{1}_{i} - Y^{2}_{i}$. Here, $Y^{1}_{i}$ represents the potential outcome observed under $T_{i}=1$, while $Y^{2}_{i}$ represents the potential outcome observed under $T_{i}=2$. Note that when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, making it a counterfactual. To address the challenge posed by these counterfactuals, the framework extends the ICE to the *average causal effect* (ACE). Researchers define the ACE as $\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]$, which represents the difference between the average of observed potential outcomes and counterfactuals.

Even when data originates from an observational study, researchers can identify the ACE from associational estimates through causal inference. They achieve this by performing statistical conditioning on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This *sufficient* set (potentially an empty set) should block all (and not open any) non-causal paths between $T$ to $Y$, thereby ensuring that the estimator for the ACE satisfies several key properties, including eliminating confounding. If such a set exists, researchers can express the ACE from associational random quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014]. Of course, the validity of the claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set. However, as @Kohler_et_al_2019 [pp. 150] noted, "statements about the real world from observed (not observational) data require assumptions (and) this is true no matter what kind of data we have". For example, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate $\tau$ if a variable $X$, such as the socio-economic status of the school, blocks all (and not open any) non-causal paths from the teaching method $T$ to the outcome $Y$.

<!-- 4. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches use rigorous mathematical notation to characterize the ACE, *not only for dichotomous but also for continuous variables* $T$, though they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, as seen in the above examples, whereas the structural approach uses do-calculus [@Pearl_2009]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to identify and estimate causal effects [@Pearl_2010].

## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}

<!-- commands for d-separation -->
\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

<!-- 1. What is the benefit of causal inference? -->
The structural approach to causal inference, however, offers a notable advantage over the potential outcomes approach. It enables researchers to formally and graphically represent a system using SCMs and DAGs [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. SCMs and DAGs function as conceptual models describing the presumed causal structure underlying the ACE. In essence, these models act as *theoretical models* that guide researchers in determining which statistical models can yield valid causal inferences, assuming the depicted causal structure of the models is correct [@McElreath_2020]. Notably, every SCM has a associated DAG [@Cinelli_et_al_2020]. @fig-IEflow provides a visual representation of the role of theoretical models in the inference process.

<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->

<!-- 2. What are the advantages of SCMs and DAGs? -->
SCMs and DAGs offer three key advantages for modeling causal structures. First, they enable researchers to represent causal relations in a non-parametric and fully interactive manner. This feature allows for feasible ACE identification strategies without specifying the data type or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Second, regardless of complexity, they can represent a wide range of causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures, facilitating their analysis by focusing on the causal assumptions associated with each building block [@McElreath_2020]. Finally, they help to formalize (causal) theoretical models into (probabilistic) statistical models for identification purposes [@McElreath_2020] [@McElreath_2020].

<!-- 3. the SCMs -->
Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks of SCMs and DAGs. The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X,Z,Y\}$, a set of *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020]. Endogenous variables are those whose causal mechanisms a researchers chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence of the variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

<!-- 4. The DAGs -->
The right panels of the figures display the complementary DAGs. A DAG consists of nodes connected by edges, where nodes represent random variables. The term *directed* means the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* signifies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. Finally, the arrows in the graphs show the expected direction of causal influences among these variables. Often, DAGs omit the exogenous variables for simplicity (the *standard* representation). However, including exogenous variables in a graph (the *magnified* representation) can be beneficial, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

::: {#fig-dags_scms1 layout-ncol=2}

::: {#fig-scm_bb1 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

SCM
:::

![DAG](images/png/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" fig-pos="H" width=50%}

<!-- ![Two unconnected nodes](images/png/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" fig-pos="H" width=35%} -->

Two unconnected nodes
:::

::: {#fig-dags_scms2 layout-ncol=2}

::: {#fig-scm_bb2 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(X,e_{Y}) \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

SCM
:::

![DAG](images/png/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" fig-pos="H" width=50%}

<!-- ![Two connected nodes or descendant](images/png/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" fig-pos="H" width=35%} -->

Two connected nodes or descendant
:::

::: {#fig-dags_scms3 layout-ncol=2}

::: {#fig-scm_bb3 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Z} \\
  e_{Z} & \dsep e_{Y}
\end{aligned}
$$

SCM
:::

![DAG](images/png/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" fig-pos="H" width=60%}
